---
title: "Markov Simulation"
output: html_document
date: "2026-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rio)
library(expm)
source("state3_after_n_trans.R")
```

# Simulation of the geometric weights 

## Simulation of run lengths given a check rate

First, a function to get a random run length given a check ("success") probability

```{r}
#get a random run length
get_run_length <- function(cr, attemps = 1000){
  
  rand <- runif(attemps)  #this is essentially generating trials
  return(which(rand <= cr)[1]) #this is finding the index (run length) of the first success
  
}

N <- 1000
cr <- 0.02

```

Now, let's simulate a $N = `r N`$ of runs with a `r cr*100` % check rate.

```{r}


simulated_run_lengths <- lapply(1:1000, function(x){
  get_run_length(cr = cr)
}) %>% unlist()
```

Here, I am writing a function that gets the actual geometric weights

```{r}
get_geometric_run_weights <- function(cr, runs = 1000){
  r <- 1:runs
  w <- cr * (1 - cr)^(r - 1)
  #w <- w / sum(w)
  return(w)
}
```


Let's check that the geometric distribution lines up with the predictions

```{r}

weights <- get_geometric_run_weights(cr)
weights <- weights*N

ggplot()+
  geom_histogram(aes(x = simulated_run_lengths), binwidth = 1)+
  geom_line(aes(x = seq_along(weights), y = weights))+
  coord_cartesian(xlim = c(0, 50))+
  theme_classic()


```


Here I'm building a function to generate a trial sequence given the switch parameters.

```{r}
gen_trial_seq <- function(N = 100, 
                          p_A2B = 0.1,
                          p_B2A = 0.1,
                          start = "A_detected"){
  
  ran <- runif(N)
  sequence <- character(N)
  sequence[1] <- start
  
  current_state <- sequence[1]
  
  for(i in 2:N){
    
    if (current_state == "A"|current_state == "A_detected"){
      if (ran[i] <= p_A2B){
        current_state <- "B"
      }
    } else if (current_state == "B"){
      if (ran[i] <= p_B2A){
        current_state <- "A"
      }
    }
    sequence[i] <- current_state
  }
  
  return(sequence)
}
```

Does the generated trial sequence line up with the simulation?

```{r}
# Simulating 1000 sequences of 100 trials. 

sequences <- list()

for (i in 1:1000){
  sequences[[i]] <- data.frame(sequence = gen_trial_seq(p_A2B = 0.1, 
                                                        p_B2A = 0.1, 
                                                        N = 100, 
                                                        start = "A_detected")) %>%
    mutate(iteration = i)
}


sequences <- bind_rows(sequences)

# calculate probabilities
sequences_agg <- sequences %>%
  group_by(iteration) %>%
  mutate(trial = 0:(n()-1)) %>%
  group_by(trial) %>%
  summarize(sim_p_A = mean(sequence == "A"),
            sim_p_A_detected = mean(sequence == "A_detected"),
            sim_p_B = mean(sequence == "B"))

# now calculate the markov states.

calculated <- data.frame(trial = 1:100) %>%
  rowwise() %>%
  mutate(calc_p_A = state3_after_n_trans(p_A2B = 0.1, 
                                         p_B2A = 0.1, 
                                         p_initial_B = 0, 
                                         p_initial_A_detected = 1, 
                                         p_initial_A = 0, 
                                         n = trial)$p_state_A,
         calc_p_A_detected = state3_after_n_trans(p_A2B = 0.1, 
                                                  p_B2A = 0.1, 
                                                  p_initial_B = 0, 
                                                  p_initial_A_detected = 1, 
                                                  p_initial_A = 0, 
                                                  n = trial)$p_state_A_detected,
         calc_p_B = state3_after_n_trans(p_A2B = 0.1, 
                                         p_B2A = 0.1, 
                                         p_initial_B = 0, 
                                         p_initial_A_detected = 1, 
                                         p_initial_A = 0, 
                                         n = trial)$p_state_B)

sequences_agg <- left_join(sequences_agg, calculated)

ggplot(sequences_agg, aes(x = trial))+
  geom_line(aes(y = calc_p_A), color = "red", lwd = 2, alpha = .2)+
  geom_line(aes(y = calc_p_A_detected), color = "green", lwd = 2, alpha = .2)+
  geom_line(aes(y = calc_p_B), color = "blue", lwd = 2, alpha = .2)+
  geom_line(aes(y = sim_p_A), color = "red")+
  geom_line(aes(y = sim_p_A_detected), color = "green")+
  geom_line(aes(y = sim_p_B), color = "blue")+
  theme_classic()

```

# Simulation 2

Here, I am actually not using geometric weights. Instead, I am simulating a bunch of agents that go through a number of trials with a given check rate, and given that check rate they might check or not. Note that this is a different approach compared to using run lengths. 

```{r}

simulate <- function(
    Win = 1,
    Loss = 10,
    RT_nc = 0.6,
    RT_c = 1.7,
    ITI = 0.1, 
    TTime = 90, 
    N_sim = 10,
    p_A2B = 0.1,
    p_B2A = 0.1,
    p_initial_A = 0,
    p_initial_B = 1,
    p_initial_A_detected = 0
){
  
  # TO DEBUG
  # Win = 1
  # Loss = 12
  # RT_nc = 0.7
  # RT_c = 1.7
  
  # Win = 1
  # Loss = 10
  # RT_nc = 1
  # RT_c = 1.5
  # ITI = 0.1
  # TTime = 90
  # N_sim = 1000
  # ITI = 0.1
  # p_A2B = 0.1
  # p_B2A = 0.1
  # p_inital_A = 0
  # p_initial_B = 1
  # p_inital_A_detected = 0
  
  #Trials to simulate:
  N_trials <- (TTime / (min(RT_nc, RT_c))) + 3 # + 3 for buffer
  
  
  # Let's generate a bunch of sequences first.
  sequences <- list()
  
  for (i in 1:N_sim){
    sequences[[i]] <- data.frame(sequence = gen_trial_seq(p_A2B = p_A2B, 
                                                          p_B2A = p_B2A, 
                                                          N = N_trials, 
                                                          start = sample(x = c("A", "A_detected", "B"), 
                                                                         size = 1, 
                                                                         prob = c(p_initial_A, p_initial_A_detected, p_initial_B)))) %>%
      mutate(iteration = i)
  }
  
  
  sequences <- bind_rows(sequences) %>%
    mutate(rand = runif(n())) # this will serve as the random number to determine binary checks. If you have a check rate of 0.1, then you compare 0.1 against the random number. Since getting a random number of 0.1 or less is 10 %, we can binarize that number by this statement: random_number <= cr
  
  cr <- seq(0.01, 1, by = 0.05)  
  
  results <- list()
  
  for(i in 1:length(cr)){
    #i = 10
    
    trials <- sequences %>% 
      mutate(Check = ifelse(rand < cr[i], 1 , 0), #binarize checking based on check rate
             RT = ifelse(Check == 1, RT_c, RT_nc) + ITI) %>% #determine check rate based on RTs and ITIs
      group_by(iteration) %>% 
      mutate(Trial = 1:n(), #enumerate trials, helpful for statistics later
             RT_cum = cumsum(RT), #cumsum of RT to help figure out at which time we have to cut off the block
             monster_id = ifelse(sequence == "A" & lag(sequence, 1, default = "B") == "B", 1, 0), #determine the monster ID, helpful to later figure out if we checked the monster or not; default "B" because we might start with state "A" (where you need to kill the monster in the very first trial) - this is a fairly unrealistic scenario that we probably won't implement, but it's good to have that possibility built in.
             monster_id = cumsum(monster_id),
             monster_id = ifelse(sequence == "B", 0, monster_id)) %>%
      group_by(iteration, monster_id) %>%
      mutate(n_checks = cumsum(Check)) %>% #this is helpful to determine whether we have checked a monster just before it turns off.
      group_by(iteration) %>%
      mutate(Rew = ifelse(sequence == "A" & lead(sequence) == "B" & n_checks == 0 & monster_id != 0, Win - Loss, Win), # if we missed a monster, where miss is defined as the monster turning off in the next trial.
             Rew_cum = cumsum(Rew)) # cumulative reward.
    
    
    iteration_summary <- trials %>%
      filter(RT_cum < TTime) %>% 
      summarize(Rew_cum = Rew_cum[n()],
                RT_M = mean(RT),
                N_trials = Trial[n()]) %>%
      mutate(CR = cr[i])
    
    trial_summary <- trials %>%
      group_by(Trial) %>%
      summarize(p_A = mean(sequence == "A"),
                p_A_detected = mean(sequence == "A_detected"),
                p_B = mean(sequence == "B"),
                Rew = mean(Rew),
                RT_M = mean(RT)) %>%
      mutate(CR = cr[i])
    
    results[[i]] <- list(iteration_summary = iteration_summary,
                         trial_summary = trial_summary)
    
  }
  
  return(results)
  
}
```


## Simulation settings and run

```{r}
# SETTINGS
Win = 1
Loss = 1
RT_nc = 0.8
RT_c = 1.9
p_A2B <- 0.1
p_B2A <- 0.1
p_initial_A_detected = 0
p_initial_A = 0 
p_initial_B = 1 
ITI = 0.1
TTime = 100
# if (start == "A_detected"){
#   p_initial_A_detected = 1
# } else  if (start == "A"){
#   p_initial_A = 1
# } else  if (start == "B"){
#   p_initial_B = 1
# }

# Simulate
results <- simulate(Win = Win, 
                    Loss = Loss, 
                    RT_nc = RT_nc,
                    RT_c = RT_c, 
                    N_sim = 1000, 
                    p_A2B = p_A2B, 
                    p_B2A = p_B2A,
                    ITI = 0.1, 
                    TTime = 90, 
                    p_initial_A = p_initial_A, 
                    p_initial_B = p_initial_B, 
                    p_initial_A_detected = p_initial_A_detected)
```

### Simulated optimality curve
Now, let's get the optimality curve.

```{r}


iteration_summaries <- lapply(results, function(x){
  x$iteration_summary
})

iteration_summaries <- bind_rows(iteration_summaries)

simulation_optimality_plt <- iteration_summaries %>% 
  group_by(CR) %>% 
  summarize(Rew_cum = mean(Rew_cum)) %>%
  ggplot(aes(x = CR, y = Rew_cum))+
  geom_line()+
  coord_cartesian(ylim = c(-100, 100))+
  labs(x = "Check Rate",
       y = "Reward",
       title = "Simulated optimality curve")+
  theme_classic()

simulation_optimality_plt

```

For comparison, also calculate the markov state probabilities

```{r}
# Also calculate the states

calculated <- data.frame(trial = 0:100) %>%
  rowwise() %>%
  mutate(calc_p_A = state3_after_n_trans(p_A2B = p_A2B, 
                                         p_B2A = p_B2A, 
                                         p_initial_B = p_initial_B, 
                                         p_initial_A_detected = p_initial_A_detected, 
                                         p_initial_A = p_initial_A, 
                                         n = trial)$p_state_A,
         calc_p_A_detected = state3_after_n_trans(p_A2B = p_A2B, 
                                                  p_B2A = p_B2A, 
                                                  p_initial_B = p_initial_B, 
                                                  p_initial_A_detected = p_initial_A_detected, 
                                                  p_initial_A = p_initial_A, 
                                                  n = trial)$p_state_A_detected,
         calc_p_B = state3_after_n_trans(p_A2B = p_A2B, 
                                         p_B2A = p_B2A, 
                                         p_initial_B = p_initial_B, 
                                         p_initial_A_detected = p_initial_A_detected, 
                                         p_initial_A = p_initial_A, 
                                         n = trial)$p_state_B)



```


### Simulated vs. calculcated trial sequence

Let's first look at the simulated trial sequence.

```{r}
trial_summaries <- lapply(results, function(x){
  x$trial_summary
})

trial_summaries <- bind_rows(trial_summaries)



trial_summaries %>% 
  group_by(Trial) %>% 
  summarize(p_A = mean(p_A),
            p_A_detected = mean(p_A_detected),
            p_B = mean(p_B),) %>%
  ggplot(aes(x = Trial))+
  geom_line(aes(y = p_A, color = "p_A"))+
  geom_line(aes(y = p_A_detected, color = "p_A_detected"))+
  geom_line(aes(y = p_B, color = "p_B"))+
  geom_line(data = calculated, aes(x = trial + 1, y = calc_p_A, color = "p_A"), lwd = 3, alpha = 0.5)+
  geom_line(data = calculated, aes(x = trial + 1, y = calc_p_A_detected, color = "p_A_detected"), lwd = 3, alpha = 0.5)+
  geom_line(data = calculated, aes(x = trial + 1, y = calc_p_B, color = "p_B"), lwd = 3, alpha = 0.5)+
  scale_color_manual(labels = c("p_A" = "A",
                                "p_B" = "B",
                                "p_A_detected" = "A detected"),
                     values = c("p_A" = "red",
                                "p_B" = "blue",
                                "p_A_detected" = "green"),
                     name = "Probability of State")+
  coord_cartesian(xlim = c(0, 20), ylim = c(0,1))+
  labs(y = "p",
       title = "simulated states")+
  theme_classic()
```



### Simulated optimality Curve by trial
How does the optimality curve change over time?

```{r}

simulated_plt_by_trial <- trial_summaries %>% 
  group_by(CR, Trial) %>% 
  summarize(Rew_per_s = Rew/RT_M,
            Rew_per_block = Rew_per_s * TTime) %>% 
  filter(Trial < 30) %>%
  ggplot(aes(x = CR, y = Rew_per_block, group = Trial, color = Trial))+
  geom_line()+
  scale_color_viridis_c()+
  labs(x = "Check Rate",
       y = "Reward per second",
       title = "optimality curve")+
  theme_classic()

simulated_plt_by_trial

```


## Comparing with calculated payoff
So why is this different to the predictions from this function?


```{r}

monster2 <- readRDS("monster2.RDS")


# Here, I am getting the basic functions for all combinations of starts

df <- data.frame(p_A2B = p_A2B,
                 p_B2A = p_B2A,
                 RT_nc = RT_nc,
                 RT_c = RT_c,
                 ITI = ITI,
                 Win = Win,
                 Loss = Loss,
                 p_initial_A = c(0, 0, 1), 
                 p_initial_A_detected = c(0, 1, 0), 
                 p_initial_B = c(1, 0, 0))

# This is just to defined three functions and their payoffs, we multiply them later with the respective probability of the state
# essentially, each of the three functions is generated under the assumption that the respective state is 100 % active. For each possible check rate, I calculate the earnings for these three functions. Later, when the probability of state A is 0.2, state B = 0.8 and state A detected = 0, the earnings are weighed by the probability of the respective state, i. e., 0.2 * (earnings from 100 * state A) + 0.8 * (earnings from state B).

fs <- df %>%
  rowwise() %>%
  mutate(label = case_when(p_initial_A == 1 ~ "start_A",
                           p_initial_A_detected == 1 ~ "start_A_detected",
                           p_initial_B == 1 ~ "start_B"),
         l = list(monster2(p_A2B = p_A2B,
                           p_B2A = p_B2A, 
                           RT_nc = RT_nc,
                           RT_c = RT_c, 
                           ITI = ITI, 
                           Win = Win, 
                           Loss = Loss, 
                           p_initial_A = p_initial_A, 
                           p_initial_A_detected = p_initial_A_detected, 
                           p_initial_B = p_initial_B, 
                           TTime = TTime) %>% 
                    as.data.frame()
         )) %>% 
  as.data.frame() %>%
  unnest(l)


fs_wide <- fs %>%
  pivot_wider(names_from = label, values_from = earnings, id_cols = probabilities)



#this is simply getting the states for different trials
runs_weight <- data.frame(trial = c(1:100, 1000),
                          p_A2B = p_A2B,
                          p_B2A = p_B2A,
                          p_initial_A = p_initial_A, 
                          p_initial_A_detected = p_initial_A_detected, 
                          p_initial_B = p_initial_B) %>%
  rowwise() %>%
  mutate(p_A = state3_after_n_trans(p_A2B = p_A2B, 
                                    p_B2A = p_B2A, 
                                    p_initial_A = p_initial_A, 
                                    p_initial_A_detected = p_initial_A_detected, 
                                    p_initial_B = p_initial_B,
                                    n = trial)$p_state_A,
         p_A_detected = state3_after_n_trans(p_A2B = p_A2B, 
                                             p_B2A = p_B2A, 
                                             p_initial_A = p_initial_A, 
                                             p_initial_A_detected = p_initial_A_detected, 
                                             p_initial_B = p_initial_B,
                                             n = trial)$p_state_A_detected,
         p_B = state3_after_n_trans(p_A2B = p_A2B, 
                                    p_B2A = p_B2A, 
                                    p_initial_A = p_initial_A, 
                                    p_initial_A_detected = p_initial_A_detected, 
                                    p_initial_B = p_initial_B,
                                    n = trial)$p_state_B) %>%
  mutate(p_miss = lag(p_A, 1, default = p_initial_A[1]) * p_A2B)

# we're here weighing 
runs_weight2 <- runs_weight %>%
  mutate(inits = list(fs_wide)) %>%
  unnest(inits) %>%
  mutate(earnings_weighted = p_A * start_A + 
           p_A_detected * start_A_detected + 
           p_B * start_B) %>%
  group_by(trial) %>%
  mutate(is_max = ifelse(earnings_weighted == max(earnings_weighted), 1, 0)) 

calculated_plt_by_trial <- ggplot(runs_weight2 %>% filter(trial != 1000), aes(x = probabilities, y = earnings_weighted, color = trial, group = trial))+
  geom_line()+
  geom_point(data = runs_weight2 %>% filter(is_max == 1), color = "black", pch = 23)+
  geom_line(data = runs_weight2 %>% filter(trial == 1000), color = "red", linetype = "21", lwd = 2)+
  scale_color_viridis_c(end = 0.9, name = "Start trial of run")+
  labs(title = "Optimality curves as a function of trial at which run started.",
       subtitle = "red: steady state",
       x = "check rate",
       y = "earnings, weighting method")+
  theme_classic()

calculated_plt_by_trial

calculated_plt_optimality <- ggplot(runs_weight2 %>% filter(trial == 1000), aes(x = probabilities, y = earnings_weighted, group = trial))+
  geom_line()+
  labs(title = "Optimality curves as a function of trial at which run started.",
       subtitle = "red: steady state",
       x = "check rate",
       y = "earnings, weighting method")+
  theme_classic()

cowplot::plot_grid(calculated_plt_optimality, simulation_optimality_plt, nrow = 1, align = "h")
cowplot::plot_grid(calculated_plt_by_trial, simulated_plt_by_trial, nrow = 1, align = "h")


```

# Simulation that allows the specification of 2 probabilities

cr_a: check rate if last time you checked state was A
cr_b: check rate if last time you checked state was B

```{r}
# 
# simulate2p <- function(
#     Win = 1,
#     Loss = 10,
#     RT_nc = 0.6,
#     RT_c = 1.7,
#     ITI = 0.1, 
#     TTime = 90, 
#     N_sim = 10,
#     p_A2B = 0.1,
#     p_B2A = 0.1,
#     p_initial_A = 0,
#     p_initial_B = 1,
#     p_initial_A_detected = 0
# ){
#   
#   # TO DEBUG
#   # Win = 1
#   # Loss = 12
#   # RT_nc = 0.7
#   # RT_c = 1.7
#   # Win = 1
#   # Loss = 10
#   # RT_nc = 1
#   # RT_c = 1.5
#   # ITI = 0.1
#   # TTime = 90
#   # N_sim = 1000
#   # ITI = 0.1
#   # p_A2B = 0.1
#   # p_B2A = 0.1
#   # p_inital_A = 0
#   # p_initial_B = 1
#   # p_inital_A_detected = 0
#   
#   N_sim = 100
#   
#   #Trials to simulate:
#   N_trials <- (TTime / (min(RT_nc, RT_c))) + 3 # + 3 for buffer
#   
#   
#   # Let's generate a bunch of sequences first.
#   sequences <- list()
#   
#   for (i in 1:N_sim){
#     sequences[[i]] <- data.frame(sequence = gen_trial_seq(p_A2B = p_A2B, 
#                                                           p_B2A = p_B2A, 
#                                                           N = N_trials, 
#                                                           start = sample(x = c("A", "A_detected", "B"), 
#                                                                          size = 1, 
#                                                                          prob = c(p_initial_A, p_initial_A_detected, p_initial_B)))) %>%
#       mutate(iteration = i)
#   }
#   
#   
#   sequences <- bind_rows(sequences) %>%
#     mutate(rand = runif(n())) # this will serve as the random number to determine binary checks. If you have a check rate of 0.1, then you compare 0.1 against the random number. Since getting a random number of 0.1 or less is 10 %, we can binarize that number by this statement: random_number <= cr
#   
#   cr_a <- seq(0.0, 1, by = 0.05)  
#   cr_b <- seq(0.0, 1, by = 0.05)  
#   
#   results <- list()
#   
#   paste0("Running ", length(cr_a)*length(cr_b), " iterations")
#   
#   current_it <- 0
#   for(i_cr_a in 1:length(cr_a)){
#     
#     for(i_cr_b in 1:length(cr_b)){
#     
#       current_it <- current_it + 1
#       
#       print(paste0("Progress:", round(current_it/(length(cr_a) * length(cr_b))), 2))
#       # i_cr_a <- 1
#       # i_cr_b <- 50
#       # trials <- sequences %>% 
#       #   group_by(iteration) %>% 
#       #   mutate(cr_a = cr_a[i_cr_a],
#       #          cr_b = cr_b[i_cr_b],
#       #          memory = sequence[1],
#       #          is_a_check = ifelse(rand <= cr_a, 1, 0),
#       #          is_b_check = ifelse(rand <= cr_b, 1, 0),
#       #          check = case_when(lag(memory, 1, default = memory[1]) == "A" & is_a_check == 1 ~ 1,
#       #                             lag(memory, 1, default = memory[1]) == "B" & is_b_check == 1 ~ 1, 
#       #                            .default =  0),
#       #          memory = case_when(check == 1 ~ sequence,
#       #                             check == 0 ~ lag(memory, 1, default = memory[1])))
#         
#         
#         
#       trials <- sequences %>%
#         group_by(iteration) %>%
#         mutate(memory = character(n()),
#                check = integer(n()),
#                cr = integer(n()),
#                cr_a = cr_a[i_cr_a],
#                cr_b = cr_b[i_cr_b],
#                new_iteration = ifelse(lag(iteration, default = 0) != iteration, 1, 0))
# 
# 
# 
# 
#       for (i in 1:nrow(trials)){
# 
#         if (trials$new_iteration[i] == 1){
#           trials$check[i] <- 0L #because we know with which the first block starts
#           trials$memory[i] <- trials$sequence[i]
#           trials$memory[i + 1] <- trials$memory[i]
# 
# 
#           if(trials$memory[i] == "A"){
#             trials$cr[i] <- cr_a[i_cr_a]
#           } else if (trials$memory[i] == "B"){
#             trials$cr[i] <- cr_b[i_cr_b]
#           }
# 
#           next
#         }
# 
#         if(trials$memory[i] == "A"){
#           trials$cr[i] <- cr_a[i_cr_a]
#         } else if (trials$memory[i] == "B"){
#           trials$cr[i] <- cr_b[i_cr_b]
#         }
# 
#         if(trials$rand[i] <= trials$cr[i]){
#           trials$check[i] <- 1L
#           if (i != nrow(trials)){
#             if (trials$new_iteration[i+1] == 0){
#               trials$memory[i + 1] <- trials$sequence[i]
#             }
#           }
#         } else {
#           trials$check[i] <- 0L
#           if (i != nrow(trials)){
#             if (trials$new_iteration[i+1] == 0){
#               trials$memory[i + 1] <- trials$memory[i]
#             }
#           }
#         }
#       }
#       
#       
#       ########
#       
#       
#       
#       trials <- trials %>%
#         mutate(RT = ifelse(check == 1, RT_c, RT_nc) + ITI) %>% #determine check rate based on RTs and ITIs
#         group_by(iteration) %>% 
#         mutate(Trial = 1:n(), #enumerate trials, helpful for statistics later
#                RT_cum = cumsum(RT), #cumsum of RT to help figure out at which time we have to cut off the block
#                monster_id = ifelse(sequence == "A" & lag(sequence, 1, default = "B") == "B", 1, 0), #determine the monster ID, helpful to later figure out if we checked the monster or not; default "B" because we might start with state "A" (where you need to kill the monster in the very first trial) - this is a fairly unrealistic scenario that we probably won't implement, but it's good to have that possibility built in.
#                monster_id = cumsum(monster_id),
#                monster_id = ifelse(sequence == "B", 0, monster_id)) %>%
#         group_by(iteration, monster_id) %>%
#         mutate(n_checks = cumsum(check)) %>% #this is helpful to determine whether we have checked a monster just before it turns off.
#         group_by(iteration) %>%
#         mutate(Rew = ifelse(sequence == "A" & lead(sequence) == "B" & n_checks == 0 & monster_id != 0, Win - Loss, Win), # if we missed a monster, where miss is defined as the monster turning off in the next trial.
#                Rew_cum = cumsum(Rew)) # cumulative reward.
#       
#       
#       iteration_summary <- trials %>%
#         filter(RT_cum < TTime) %>% 
#         summarize(Rew_cum = Rew_cum[n()],
#                   RT_M = mean(RT),
#                   N_trials = Trial[n()]) %>%
#         mutate(cr_a = cr_a[i_cr_a],
#                cr_b = cr_b[i_cr_b])
#       
#       trial_summary <- trials %>%
#         group_by(Trial) %>%
#         summarize(p_A = mean(sequence == "A"),
#                   p_A_detected = mean(sequence == "A_detected"),
#                   p_B = mean(sequence == "B"),
#                   Rew = mean(Rew),
#                   RT_M = mean(RT)) %>%
#         mutate(cr_a = cr_a[i_cr_a],
#                cr_b = cr_a[i_cr_b])
#       
#       results[[length(results) + 1]] <- list(iteration_summary = iteration_summary,
#                                              trial_summary = trial_summary)
#       
#     }
#   }
#   
#   return(results)
#   
# }
```



```{r}

# library(Rcpp)
# sourceCpp("memory_loop.cpp")

saveRDS(gen_trial_seq, "gen_trial_seq.RDS")

simulate2p <- function(
    Win = 1,
    Loss = 10,
    RT_nc = 0.6,
    RT_c = 1.7,
    ITI = 0.1, 
    TTime = 90, 
    N_sim = 10,
    p_A2B = 0.1,
    p_B2A = 0.1,
    p_initial_A = 0,
    p_initial_B = 1,
    p_initial_A_detected = 0
){
  
  print(getwd())
  library(Rcpp)
  sourceCpp("memory_loop.cpp")
  gen_trial_seq <- readRDS("gen_trial_seq.RDS")
  # TO DEBUG
  # Win = 1
  # Loss = 12
  # RT_nc = 0.7
  # RT_c = 1.7
  # Win = 1
  # Loss = 10
  # RT_nc = 1
  # RT_c = 1.5
  # ITI = 0.1
  # TTime = 90
  # N_sim = 1000
  # ITI = 0.1
  # p_A2B = 0.1
  # p_B2A = 0.1
  # p_inital_A = 0
  # p_initial_B = 1
  # p_inital_A_detected = 0
  
  
  
  #Trials to simulate:
  N_trials <- (TTime / (min(RT_nc, RT_c))) + 3 # + 3 for buffer
  
  
  # Let's generate a bunch of sequences first.
  sequences <- list()
  
  for (i in 1:N_sim){
    sequences[[i]] <- data.frame(sequence = gen_trial_seq(p_A2B = p_A2B, 
                                                          p_B2A = p_B2A, 
                                                          N = N_trials, 
                                                          start = sample(x = c("A", "A_detected", "B"), 
                                                                         size = 1, 
                                                                         prob = c(p_initial_A, p_initial_A_detected, p_initial_B)))) %>%
      mutate(iteration = i)
  }
  
  
  sequences <- bind_rows(sequences) %>%
    mutate(rand = runif(n()))  %>%# this will serve as the random number to determine binary checks. If you have a check rate of 0.1, then you compare 0.1 against the random number. Since getting a random number of 0.1 or less is 10 %, we can binarize that number by this statement: random_number <= cr
  group_by(iteration) %>%
  mutate(
    Trial = 1:n(),
    new_iteration = ifelse(lag(iteration, default = 0) != iteration, 1L, 0L),
    monster_id = ifelse(sequence == "A" & lag(sequence, 1, default = "B") == "B", 1, 0),
    monster_id = cumsum(monster_id),
    monster_id = ifelse(sequence == "B", 0, monster_id)
  ) %>%
  ungroup()
  
  cr_a <- seq(0.0, 1, by = 0.05)  
  cr_b <- seq(0.0, 1, by = 0.05)  
  
  results <- list()
  
  paste0("Running ", length(cr_a)*length(cr_b), " iterations")
  
  current_it <- 0
  for(i_cr_a in 1:length(cr_a)){
    
    for(i_cr_b in 1:length(cr_b)){
    
      # i_cr_a <- 10
      # i_cr_a <- 15
      
      current_it <- current_it + 1
      
      if (current_it %% 100 == 0){
        print(paste0("Progress: ", round((current_it/(length(cr_a) * length(cr_b)))*100, 2), "%"))
      }

      
      
      result <- compute_memory_check(sequences$sequence, 
                                     sequences$rand,
                                     cr_a[i_cr_a], 
                                     cr_b[i_cr_b],
                                     sequences$new_iteration)
      
      
      ########
      
      
      
      trials <- sequences %>%
      mutate(memory = result$memory,
             check = result$check,
             RT = ifelse(check == 1, RT_c, RT_nc) + ITI) %>% #determine check rate based on RTs and ITIs
        group_by(iteration) %>% 
        mutate(Trial = 1:n(), #enumerate trials, helpful for statistics later
               RT_cum = cumsum(RT), #cumsum of RT to help figure out at which time we have to cut off the block
               monster_id = ifelse(sequence == "A" & lag(sequence, 1, default = "B") == "B", 1, 0), #determine the monster ID, helpful to later figure out if we checked the monster or not; default "B" because we might start with state "A" (where you need to kill the monster in the very first trial) - this is a fairly unrealistic scenario that we probably won't implement, but it's good to have that possibility built in.
               monster_id = cumsum(monster_id),
               monster_id = ifelse(sequence == "B", 0, monster_id)) %>%
        group_by(iteration, monster_id) %>%
        mutate(n_checks = cumsum(check)) %>% #this is helpful to determine whether we have checked a monster just before it turns off.
        group_by(iteration) %>%
        mutate(Rew = ifelse(sequence == "A" & lead(sequence) == "B" & n_checks == 0 & monster_id != 0, Win - Loss, Win), # if we missed a monster, where miss is defined as the monster turning off in the next trial.
               Rew_cum = cumsum(Rew)) # cumulative reward.
      
      
      iteration_summary <- trials %>%
        filter(RT_cum < TTime) %>% 
        summarize(Rew_cum = Rew_cum[n()],
                  RT_M = mean(RT),
                  N_trials = Trial[n()]) %>%
        mutate(cr_a = cr_a[i_cr_a],
               cr_b = cr_b[i_cr_b])
      
      trial_summary <- trials %>%
        group_by(Trial) %>%
        summarize(p_A = mean(sequence == "A"),
                  p_A_detected = mean(sequence == "A_detected"),
                  p_B = mean(sequence == "B"),
                  Rew = mean(Rew),
                  RT_M = mean(RT)) %>%
        mutate(cr_a = cr_a[i_cr_a],
               cr_b = cr_a[i_cr_b])
      
      results[[length(results) + 1]] <- list(iteration_summary = iteration_summary,
                                             trial_summary = trial_summary)
      
    }
  }
  
  return(results)
  
}

saveRDS(simulate2p, "simulate2p.RDS")


```


```{r}

output <- simulate2p(
    Win = 1,
    Loss = 50,
    RT_nc = 0.6,
    RT_c = 1.7,
    ITI = 0.1, 
    TTime = 90, 
    N_sim = 100,
    p_A2B = 0.1,
    p_B2A = 0.1,
    p_initial_A = 0,
    p_initial_B = 1,
    p_initial_A_detected = 0
)
  
  


iteration_summaries <- lapply(output, function(y){
  y$iteration_summary
}) %>% bind_rows %>%
  group_by(cr_a, cr_b) %>%
  summarize(Rew_cum = mean(Rew_cum))


ggplot(iteration_summaries, aes(x = cr_a, y = Rew_cum, color = cr_b, group = cr_b))+
  geom_line()+
  scale_color_viridis_c()+
  theme_classic()


ggplot(iteration_summaries, aes(x = cr_a, y = cr_b, fill = Rew_cum, group = Rew_cum))+
  geom_tile()+
  scale_fill_viridis_c()




```





